The effect of teachers reassigning students to new Cognitive Tutor sections

### 2020-03-12 paper.Rnw
submission

\numberofauthors{2} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Adam C Sales\\
       \affaddr{University of Texas College of Education}\\
       \affaddr{Austin, TX, USA}\\
       \email{asales@utexas.edu}
% 2nd. author
\alignauthor
John F Pane\\
       \affaddr{RAND Corporation}\\
       \affaddr{Pittsburgh, PA, USA}\\
       \email{jpane@rand.org}
}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
The design of the Cognitive Tutor Algebra I (CTA1) intelligent tutoring system assumes that students work through sections of material following a pre-specified order, and only move on from one section to the next after mastering the first section’s skills. However, the software gives teachers the ability to override that structure, by reassigning students to different sections of the curriculum. Which students get reassigned? Does reassignment hurt student learning? Does it help? This paper used data from the treatment arm of a large effectiveness study of the CTA1 curriculum to estimate the effects of reassignment on students’ scores on an Algebra I posttest. Since reassignment is not randomized, we used a multilevel propensity score matching design, along with assessments of sensitivity to bias from unmeasured confounding, to estimate the effects of reassignment. We found that reassignment reduces posttest scores by roughly 0.15 standard deviations—about three-quarters of the overall CTA1 treatment effect—that unmeasured confounding is unlikely to completely explain this observed effect, but that the effect of reassignment may vary widely between classrooms.
\end{abstract}

@@

\section{Introduction}
\section{The Cognitive Tutor, Mastery Learning and Reassignment}\label{sec:background}
\section{Data: the RAND CTA1 Effectiveness Study}\label{sec:data}
In the years 2007--2010, the RAND Corporation conducted a randomized
study to test the effectiveness of the CTA1 curriculum relative to
business as usual.
The study tested CTA1 under authentic, natural conditions---that is,
oversight and support of CTA1's use was the same as it would have been
outside of an RCT.
The study population consisted of over 25,000 students in 73 high schools and 74 middle
schools located in 52 diverse school districts in seven states.
Students in Algebra I classrooms in participating schools took an algebra I
pretest and a posttest, both from the CTB/McGraw-Hill Acuity series.
The pretest was the Algebra Readiness Exam, a 40-item multiple-choice
exam testing students' algebra I prerequisite skills.
The posttest was the Algebra Proficiency Exam, a 32-item
multiple-choice exam testing algebra I skills including solving
equations for an unknown, graphing linear and quadratic functions,
calculating complex algebraic expressions and other skills.
Data from both exams were scored with a three-parameter item
response theory (IRT) model.

@@

Unfortunately, log data was not available for every student in the
treatment arm of the study, primarily for two reasons: some students
in CTA1 schools nevertheless did not use the tutor, and some students
used the tutor but their log data was irretreavable or could not be
reliably linked to posttest scores and covariates.
This study omitted schools in which data was missing for over 20\% of
students in either year, leaving \Sexpr{pn(nschool)} schools.
Among the students at these schools, we omitted 164 had no log data, and 242
either only worked---but did not complete---one section or had no
section completion data for some other reason.
A total of \Sexpr{pn(nstud)} students in \Sexpr{pn(nclass)} classrooms
remained in the analysis sample, roughly \Sexpr{percentKept}\% of the
full treatment group.

@@

Let $\bm{x}_i$ denote a vector of covariates for subject $i$.
These include pretest scores, special education, gifted, and English
language learner (ELL) status,
race/ethnicity (white, black, Latinx\footnote{For the sake of
  parsimony, these categories were
  collapsed from a larger set in the original dataset, so that
  \Sexpr{nInd} American Indian/Alaskan Native students were
  catagorized as Latinx, \Sexpr{nAsn} Asian/Pacific Islander students
  and \Sexpr{nRaceNA} students with missing data were categorized as white, and \Sexpr{nOth} Other/Multiracial
  students were categorized as black.}),
received free or reduced-price lunch (FRL).
Let $Class_i$ be $i$'s classroom; since reassignment occured within
classrooms, $Class$ is a covariate as well.
If reassignment were randomly assigned, the (theoretical) distribution
of $\bm{x}$ and $Class$
would be equal between reassigned and not-reassigned
students---$\bm{x}$ and $Class$ would be balanced.
Our strategy will be to construct a randomization scheme in which
$\bm{x}$, and, to the extent possible, $Class$ are balanced, and
conduct inference under that randomization scheme.

@@

The first condition for propensity score matching is that there is some randomness in the treatment assignment:
\begin{equation}\label{eq:commonSupport}
0<e_i(\bm{x}_i,Class_i)<1 \text{ for all }i.
\end{equation}
When \eqref{eq:commonSupport} fails for a subset of the analysis sample, common practice is to drop that subset and estimate average effects for the remainder of the analysis sample, i.e. the subset for which \eqref{eq:commonSupport} holds; this subset is referred to as the ``region of common support" \cite{caliendo2008some} \cite{shadish2010primer}.
In this study, including $Class$ among the covariates leads to violations of \eqref{eq:commonSupport}.
Of the \Sexpr{nclassTot}
classrooms over the two years of the study, \Sexpr{nclass0} contained no
reassigned students, and in \Sexpr{nclass1} classrooms every student
was reassigned at least once.
In this subset of the data, including \Sexpr{round(pstud10*100)}\% of
students, $Pr(Z=1|Class)=0$ or 1.
Our solution is to analyze the data in two parts.
In the first part, we drop classrooms in which no one or everyone was reassigned, and only estimate effects for students in classrooms with some reassignment variance, a student-level analysis.
In the second part, we conduct a classroom-level analysis.
For for classrooms $j=1,...,J$, let $\bar{Z}_j=1$ if everyone in classroom $j$ was reassigned and $\bar{Z}=0$ if no one was reassigned.
Then we estimate propensity scores $e(\bm{\bar{x}}_j)=Pr(\bar{Z}=1|\bm{\bar{Z}})$, where $\bm{\bar{x}}$ is a vector of covariates aggregated to the classroom level, and match $\bar{Z}=1$ classrooms to $\bar{Z}=0$ classrooms with similar propensity scores.
Finally, we estimate effects for these classrooms by modeling $Z$ as assigned at the classroom level, as in a group randomized trial.
We will present the student-level and classroom-level effect estimates separately as well as in a combined model.

The second condition for propensity score matching is that there are no unmeasured confounders:
\begin{equation}\label{eq:ignore1}
(y^1,y^0)\independent Z|\bm{x},Class
\end{equation}
for the student-level analysis and
\begin{equation*}
(y^1,y^0)\independent \bar{Z}|\bm{\bar{x}}
\end{equation*}
for the classroom-level analysis.
Assumption \eqref{eq:ignore1} is well known as the achilles heel of causal inference outside of RCTs.
\eqref{eq:ignore1} is untestable; its believability depends on what is understood about the process that underlies treatment assignment $Z$, and what covariates are available for control.
In our case, reassignment is poorly understood, and appears highly idiosyncratic \cite{descriptivePaper}.
Fortunately, our study includes a pretest measure, and observational studies controlling for pretest scores tend to perform well, and repilicate experimental estimates \cite{cook2008three} \cite{cook2009bias}.
Section \ref{sec:sensitivity} discusses a sensitivity analysis that
relaxes \label{eq:ignore1} and assumes reasonable levels of unmeasured
confounding.

Our attitude towards propensity score matching is agnostic.
If the propensity score models in the following section were
approximately corrrect, and yielded good estimates of the true
propensity scores, then the theory underlying propensity score
adjustment holds.
If not, the process of propensity score matching may still result in a
set of matched reassigned and not reassigned students that, on
average, resemble each other on all measured covariates.
In other words, the (mis)estimated propensity scores $\hat{e}$ may still be
approximate ``balancing'' scores, satisfying
\begin{equation}\label{eq:balancing}
\bm{x}\independent Z|\hat{e}.
\end{equation}
Causal inference based on comparisons within these matched sets will
still be plausible; indeed, \cite{rosenbaum1983central} showed that in order to estimate
average treatment effects, it is sufficient to condition on a
balancing score, rather than the propensity score itself.

@@

All data analysis was done in \texttt{R} \cite{rcite} using the
\texttt{tidyverse} suite of packages \cite{tidyverse} for data
manipuation, plotting, and other tasks.
This document was produced dynamically with \texttt{knitr}
\cite{knitr}.
Source code is available at \url{www.github.com/adamSales/cpEffect}.

\section{Propensity Scores: Who Gets Reassigned?}\label{sec:propensityScores}
\subsection{Student-Level}
We use multilevel logistic regression \cite{gelmanHill} to estimate student level propensity scores.
The multilevel regression accounts for the nesting of students within classrooms, classrooms within teachers, and teachers within schools.
In constructing the model, we give special consideration to the role of pretest scores, a proxy for student mathematical ability at the beginning of the school year, in predicting reassignment.
First, we decompose pretest scores into student- and classroom-level components.
If $w_i$ is student $i$'s pretest score, let $w_i=\bar{w}_{j[i]}+\tilde{w}_i$, where $\bar{w}_{j[i]}$ is the average pretest score in $i$'s classroom $j[i]$, and $\tilde{w}_i$ is the difference between $i$'s pretest score and the classroom mean.
This decomposition was motivated by the possibility that reassignment patterns may differ between high- and low-achieving classrooms, and that a teacher's decision to reassign a student depends on the student's ability relative to the classroom than his or her absolute ability.
Second, we modeled the effect of $\tilde{w}$ on $Z$ as linear in the logit scale, but allowed the slope to vary by classroom.
This was motivated by the possibilty that some teachers use reassignment to help struggling students catch up to their peers, so lower $\tilde{w}$ would predict $Z$, and other teachers use it to help high-achievers skip sections related to basic skills, so higher $\tilde{w}$ would predict $Z$.
We also considered models incorporating non-linear effects of
$\tilde{w}$, via natural splines \cite{hastie2017generalized} but
found no evidence that the non-lineararity improved the model fit.
We fit the model using the \texttt{lme4} package in \texttt{R}
\cite{lme4} \cite{rcite}.

@@

\subsection{Classroom-Level}
Using data from classrooms in which either all students or no students
were reassigned, we estimated classroom-level propensity scores
$Pr(\bar{Z}=1|\bm{\bar{x}})$.
To do so, we used a single-level logistic regression model:
\begin{equation}\label{eq:propmod2}
\begin{split}
Pr(&\bar{Z}_j=1|\bm{\bar{x}}_j)=\\
&logit^{-1}( \gamma_{0state[j]}+\gamma_1 \bar{w}_{j}+\\
&\gamma_2 \%Freshman_j+\gamma_{3} \%FRL_j+\\
&\gamma_{4}year_j+\gamma_5Nstud_j)
\end{split}
\end{equation}
where $\%Freshman_j$ and $\% FRL_j$ denote the percentages of
classroom $j$ that are 9th graders or receive free or reduced-price
lunches, and $Nstud_j$ is the number of students in classroom $j$ in
our dataset.
Model \eqref{eq:propmod2} was simpler than the student-level version,
\eqref{eq:propmod1}, both because it used classroom-aggregated data,
and because the sample size was smaller ($n=\Sexpr{nclass1+nclass0}$
classrooms versus $n=nStud1$ students), requiring greater parsimony.
As described in Section \ref{sec:psm}, model \eqref{eq:propmod2} was
selected in order to acheive covariate balance \eqref{eq:balance}.

\begin{figure}\label{fig:coefplot2}
<<coefplot2,fig.width=3,fig.height=2>>=

pdat2 <- psmod2coef%>%
  as.data.frame()%>%
  rownames_to_column('Covariate')%>%
  filter(Covariate%in%c('nstud','mpretest','raceBlack','raceHispanic','sexM','grade9','spec_speced', 'spec_gifted','spec_esl','frl','frlMISTRUE','year'))%>%
  mutate(
    Covariate= map_chr(Covariate,
      ~switch(.,
        mpretest='wbar',#"$\\bar{w}$",
        pretestC='wtilde',#"$\\tilde{w}$",
        raceWhite='White',
        raceBlack='Black',
        raceHispanic='Latinx',
        sexF='F',
        sexM='Male',
        "grade10+"="10+",
        grade9="%Freshman",
        spec_speced="Special Ed.",
        spec_gifted="Gifted",
        spec_esl="ESL",
        frl="%FRL",
        frlMISTRUE="FRL Missing",
        year="Year",
        nstud="Nstud")
    ),
    Covariate=factor(Covariate,levels=rev(unique(Covariate))),
    ymax=Estimate+2*`Std. Error`,
    ymin=Estimate-2*`Std. Error`
  )

ggplot(pdat2,aes(Covariate,Estimate,ymin=ymin,ymax=ymax))+
      geom_point(size=3)+
          xlab(NULL)+ylab(NULL)+
  geom_errorbar(width=0,size=2)+geom_hline(yintercept=0)+
  scale_x_discrete(labels=c(levels(pdat2$Covariate)[1:(nlevels(pdat2$Covariate)-1)],bquote(bar(w))))+
  theme(text = element_text(size=10))+
  coord_flip()

@
\caption{Estimated coefficients and 95\% confidence intervals from
  model \eqref{eq:propmod2}.}
\label{fig:propmodel2}
\end{figure}

Figure \ref{fig:propmodel2} shows estimates and 95\% confidence
intervals for covariate coefficients from \eqref{eq:propmodel2}.
Similar to the results from the student-level analysis, reassignment
is more prevalent in classrooms with lower average pretest scores and
in the second year of implementation, though \eqref{eq:propmodel2}
cannot rule out null, or very small opposite-signed effects in either case.
Teachers in classrooms with a greater proportion of FRL students were less likely
to use reassignment.

@@

\section{Matching and Covariate Balance}
\subsection{Student Level Match}\label{sec:match}
We construct a student-level match based on propensity scores on the
log-odds scale, i.e. $log(\hat{e}/(1-\hat{e})$.
Instead of a pair-matching design, which would necessitate discarding
non-reassigned students who would make good matched comparisons, we
use a restricted full match design \cite{hansen2004}.
In this design, the numbers of reassigned and not-reassigned students
in each matched set is allowed to vary, so that in some cases several
reassigned students may be matched with a single non-reassigned
studnet, and vice-versa.
We use the \texttt{R} package \texttt{optmatch} \cite{optmatch} to
choose the matched sets optimally.
The \texttt{fullmatch()} routine takes a matrix of discrepencies
(e.g. differences in propensity scores) between treatment and control
subjects, and arranges them into matched sets so that the sum of
absolute discrepencies between matched subjects is minimized.

As described at the end of Section \ref{sec:psm}, the post-test scores
played no role in this process.
Hence, we were able to iteratively match students, check covariate
balance, modify the propensity score model and/or the matching routine
if necessary, and repeat until adaquate balance was achieved.
Here we present the final match; a record of attempts is available on
the first author's github site.

\input{tables/balance1.tex}
\subsection{Classroom Level Match}
\subsection{Effect Estimates}
\section{The Effect of Reassignment}\label{sec:effects}
\section{What About Unobserved Confounders?}\label{sec:sensitivity}
\section{Treatment Effect Heterogeneity}\label{sec:heterogeneity}
\section{Discussion}\label{sec:discussion}

<<include=FALSE,cache=FALSE>>=
detach(dataInfo)
@
\section{Acknowledasdfasdfgments}
This section is optional; it is a location for you
to acknowledge grants, funding, editing assistance and
what have you.  In the present case, for example, the
authors would like to thank Gerald Murray of ACM for
his help in codifying this \textit{Author's Guide}
and the \textbf{.cls} and \textbf{.tex} files that it describes.

>>>>

%\numberofauthors{2} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
%\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
% \alignauthor
% Adam C Sales\\
%        \affaddr{University of Texas College of Education}\\
%        \affaddr{Austin, TX, USA}\\
%        \email{asales@utexas.edu}
% % 2nd. author
% \alignauthor
% John F Pane\\
%        \affaddr{RAND Corporation}\\
%        \affaddr{Pittsburgh, PA, USA}\\
%        \email{jpane@rand.org}
% }
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
The design of the Cognitive Tutor Algebra I (CTA1) intelligent
tutoring system assumes that students work through sections of
material following a pre-specified order, and only move on from one
section to the next after mastering the first section’s
skills. However, the software gives teachers the ability to override
that structure, by reassigning students to different sections of the
curriculum. Which students get reassigned? Does reassignment hurt
student learning? Does it help? This paper used data from the
treatment arm of a large effectiveness study of the CTA1 curriculum to
estimate the effects of reassignment on students’ scores on an Algebra
I posttest. Since reassignment is not randomized, we used a multilevel
propensity score matching design, along with assessments of
sensitivity to bias from unmeasured confounding, to estimate the
effects of reassignment. We found that reassignment reduces posttest
scores by roughly 0.2 standard deviations—about the same as the overall CTA1 treatment effect—that unmeasured confounding is unlikely to completely explain this observed effect, but that the effect of reassignment may vary widely between classrooms.
\end{abstract}

@@

\section{Introduction}
Two closely related pillars of intelligent tutoring systems are
sequencing and mastery learning.
It has long been obvious that the sequence in which students learn
different topics is an important component of a curriculum, due to
prerequisites---for instance, students must master arithmetic in order
to learn how to solve algebraic equations.
A related example is scaffolding, in which
learners gradually achieve independence over a sequence of problems;
scaffolding ``consists essentially of the adult 'controling' those
elements of the task that are initially beyond the learner's capacity,
thus permitting him to concentrate upon and complete only those
elements that are within his range of competence''
\cite{wood1976role}.
However, measuring the effects of sequencing \cite{ritter2007order}
\cite{doroudi2016sequence} and determining prerequisites or optimal
sequences \cite{scheines2014discovering} \cite{vuong2011method} \cite{koedinger2002toward}
remains an active area of research.

By ``mastery learning,'' we mean the idea that students should
``progress through  topics  as  they  master  them,''
\cite{ritter2013predicting} as opposed to at a fixed pace.
This typically results in students within the same classroom working
on different parts of a curriculum at the same time.

The Cognitive Tutor Algebra I (CTA1) system
\cite{corbett2001cognitive} includes both features.
A particular Algebra I curriculum is programmed into the software, so
that students, if left alone, will encounter topics in a specific,
intentional sequence.
Mastery learning governs how they progress from one section to the
next: an underlying knowledge tracing model estimates the probability
students have mastered a set of pre-defined skills as they work
through problems that incorporate those skills.
Students ideally progress from one section to the next only after
demonstrating mastery on the previous section's skills.

Mastery learning does not always proceed this way in the CTA1
software.
After a student has worked a certain, pre-specified number of problems
in a particular section, he or she is automatically promoted to the
next section, even if he or she has not mastered its skills
\cite{aoas}.
Teachers can also reassign students working on one section to work on
an entirely different section.
If a teacher reassigns a student to a section other than the next one
in the sequence, reassignment violates the intended sequencing as well
as mastery learning.

There are a number of reasons teachers may want to meddle in the
automatic progress of students through a curriculum \cite{descriptivePaper}.
If a teacher observes an advanced student spending time on basic
skills, the teacher may move the student to more advanced sections.
If certain skills will be on a standardized test, and a teacher wants
all students to have had exposure to those skills before the test, the
teacher may reassign all of his or her students to work on a section
covering those skills.
If a teacher notices a student falling behind his or her peers in the
classroom, the teacher may choose to reassign the student to the
section that the rest of the class is working on, even if the student
has not demonstrated mastery on prerequisite skills (at least, within
the tutor).
If a teacher disagrees with the method a certain CTA1 section employs
in teaching an Algebra topic, the teacher may reassign students out of
that section, perhaps to the next unit or section in the curriculum.

It is unclear whether reassignment benefits students.
On the one hand, it violates the design principles of the software.
On the other hand, it allows teachers flexibility to teach the
material as they see fit, and use the tutor to meet the particular
needs of their classrooms.

This paper uses data from a large randomized trial of the CTA1
curriculum to estimate the effect of reassignment.
Unfortunately for our purposes, reassignment itself was not
randomized---the study was designed to estimate CTA1's effectiveness,
so access to the tutor was randomized instead.
Still, log data from study participants includes data on how often
each student was reassigned from one section to another, and posttests
measure their algebra skills at the end of the study.
For those reasons, this data provides a rare opportunity to measure
the effect of reassignment, and, by extension, the (joint)  importance
of topic sequencing and mastery learning.

The following section gives background on the effectiveness trial
and describe the data we will use for the study.
Section \ref{sec:psm} describes propensity score matching, the
method we employ. Section \ref{sec:propensityScores} describes the
propensity score models, which in turn describe characteristics of
students who are reassigned. Section \ref{sec:match} describes the
matching algorithm and covariate balance. Section \ref{sec:effects}
gives our main results on the effects of reassignment, including
sensitivity analysis to confounding from unmeasured covariates and
between-classroom effect heterogeneity. Section \ref{sec:discussion} concludes.

\section{Data: the RAND CTA1 Effectiveness Study}\label{sec:data}
In the years 2007--2010, the RAND Corporation conducted a randomized
study to test the effectiveness of the CTA1 curriculum relative to
business as usual.
The study tested CTA1 under authentic, natural conditions---that is,
oversight and support of CTA1's use was the same as it would have been
outside of an RCT.
The study population consisted of over 25,000 students in 73 high schools and 74 middle
schools located in 52 diverse school districts in seven states.
Students in Algebra I classrooms in participating schools took an algebra I
pretest and a posttest, both from the CTB/McGraw-Hill Acuity series.
The pretest was the Algebra Readiness Exam, a 40-item multiple-choice
exam testing students' algebra I prerequisite skills.
The posttest was the Algebra Proficiency Exam, a 32-item
multiple-choice exam testing algebra I skills including solving
equations for an unknown, graphing linear and quadratic functions,
calculating complex algebraic expressions and other skills.
Data from both exams were scored with a three-parameter item
response theory (IRT) model.

@@

Unfortunately, log data was not available for every student in the
treatment arm of the study, primarily for two reasons: some students
in CTA1 schools nevertheless did not use the tutor, and some students
used the tutor but their log data was irretrievable or could not be
reliably linked to posttest scores and covariates.
This study omitted schools in which data was missing for over 20\% of
students in either year, leaving \Sexpr{pn(nschool)} schools.
Among the students at these schools, we omitted 164 had no log data, and 242
either only worked---but did not complete---one section or had no
section completion data for some other reason.
A total of \Sexpr{pn(nstud)} students in \Sexpr{pn(nclass)} classrooms
remained in the analysis sample, roughly \Sexpr{percentKept}\% of the
full treatment group.

@@

Let $\bm{x}_i$ denote a vector of covariates for subject $i$.
These include pretest scores, special education, gifted, and English
language learner (ELL) status,
race/ethnicity (white, black, Latinx\footnote{For the sake of
  parsimony, these categories were
  collapsed from a larger set in the original dataset, so that
  \Sexpr{nInd} American Indian/Alaskan Native students were
  categorized as Latinx, \Sexpr{nAsn} Asian/Pacific Islander students
  and \Sexpr{nRaceNA} students with missing data were categorized as white, and \Sexpr{nOth} Other/Multiracial
  students were categorized as black.}),
received free or reduced-price lunch (FRL).
Let $Class_i$ be $i$'s classroom; since reassignment occurred within
classrooms, $Class$ is a covariate as well.
If reassignment were randomly assigned, the (theoretical) distribution
of $\bm{x}$ and $Class$
would be equal between reassigned and not-reassigned
students---$\bm{x}$ and $Class$ would be balanced.
Our strategy will be to construct a randomization scheme in which
$\bm{x}$, and, to the extent possible, $Class$ are balanced, and
conduct inference under that randomization scheme.

@@

The first condition for propensity score matching is that there is some randomness in the treatment assignment:
\begin{equation}\label{eq:commonSupport}
0<e_i(\bm{x}_i,Class_i)<1 \text{ for all }i.
\end{equation}
When \eqref{eq:commonSupport} fails for a subset of the analysis sample, common practice is to drop that subset and estimate average effects for the remainder of the analysis sample, i.e. the subset for which \eqref{eq:commonSupport} holds; this subset is referred to as the ``region of common support" \cite{caliendo2008some} \cite{shadish2010primer}.
In this study, including $Class$ among the covariates leads to violations of \eqref{eq:commonSupport}.
Of the \Sexpr{nclassTot}
classrooms over the two years of the study, \Sexpr{nclass0} contained no
reassigned students, and in \Sexpr{nclass1} classrooms every student
was reassigned at least once.
In this subset of the data, including \Sexpr{round(pstud10*100)}\% of
students, $Pr(Z=1|Class)=0$ or 1.
Our solution is to drop classrooms in which no one or everyone was
reassigned, and only estimate effects for students in classrooms with
some reassignment variance, a student-level analysis.

We attempted a parallel classroom-level analysis, in which we matched
classrooms in which all students were reassigned to classrooms in
which no one was.
However, we were unable to construct a match with adequate covariate
balance (there were few no-reassigned classrooms with similar mean
pretest scores to the all-reassigned classrooms that were of similar
sizes).
For that reason, we dropped the classroom-level analysis.
% In the second part, we conduct a classroom-level analysis.
% For for classrooms $j=1,...,J$, let $\bar{Z}_j=1$ if everyone in classroom $j$ was reassigned and $\bar{Z}=0$ if no one was reassigned.
% Then we estimate propensity scores $e(\bm{\bar{x}}_j)=Pr(\bar{Z}=1|\bm{\bar{Z}})$, where $\bm{\bar{x}}$ is a vector of covariates aggregated to the classroom level, and match $\bar{Z}=1$ classrooms to $\bar{Z}=0$ classrooms with similar propensity scores.
% Finally, we estimate effects for these classrooms by modeling $Z$ as assigned at the classroom level, as in a group randomized trial.
% We will present the student-level and classroom-level effect estimates separately as well as in a combined model.

The second condition for propensity score matching is that there are no unmeasured confounders:
\begin{equation}\label{eq:ignore1}
(y^1,y^0)\independent Z|\bm{x},Class
\end{equation}
% for the student-level analysis and
% \begin{equation*}
% (y^1,y^0)\independent \bar{Z}|\bm{\bar{x}}
% \end{equation*}
% for the classroom-level analysis.
Assumption \eqref{eq:ignore1} is well known as the Achilles heel of causal inference outside of RCTs.
\eqref{eq:ignore1} is untestable; its believability depends on what is understood about the process that underlies treatment assignment $Z$, and what covariates are available for control.
In our case, reassignment is poorly understood, and appears highly idiosyncratic \cite{descriptivePaper}.
Fortunately, our study includes a pretest measure, and observational studies controlling for pretest scores tend to perform well, and replicate experimental estimates \cite{cook2008three} \cite{cook2009bias}.
Section \ref{sec:sensitivity} discusses a sensitivity analysis that
relaxes \label{eq:ignore1} and assumes reasonable levels of unmeasured
confounding.

Our attitude towards propensity score matching is agnostic.
If the propensity score models in the following section were
approximately correct, and yielded good estimates of the true
propensity scores, then the theory underlying propensity score
adjustment holds.
If not, the process of propensity score matching may still result in a
set of matched reassigned and not reassigned students that, on
average, resemble each other on all measured covariates.
In other words, the (mis)estimated propensity scores $\hat{e}$ may still be
approximate ``balancing'' scores, satisfying
\begin{equation}\label{eq:balancing}
\bm{x}\independent Z|\hat{e}.
\end{equation}
Causal inference based on comparisons within these matched sets will
still be plausible; indeed, \cite{rosenbaum1983central} showed that in order to estimate
average treatment effects, it is sufficient to condition on a
balancing score, rather than the propensity score itself.

@@

All data analysis was done in \texttt{R} \cite{rcite} using the
\texttt{tidyverse} suite of packages \cite{tidyverse} for data
manipulation, plotting, and other tasks.
This document was produced dynamically with \texttt{knitr}
\cite{knitr}.
Source code is available at www.github.com/[Redacted].

\section{Propensity Scores: Who Gets Reassigned?}\label{sec:propensityScores}
We use multilevel logistic regression \cite{gelmanHill} to estimate student level propensity scores.
The multilevel regression accounts for the nesting of students within classrooms, classrooms within teachers, and teachers within schools.
In constructing the model, we give special consideration to the role of pretest scores, a proxy for student mathematical ability at the beginning of the school year, in predicting reassignment.
First, we decompose pretest scores into student- and classroom-level components.
If $w_i$ is student $i$'s pretest score, let $w_i=\bar{w}_{j[i]}+\tilde{w}_i$, where $\bar{w}_{j[i]}$ is the average pretest score in $i$'s classroom $j[i]$, and $\tilde{w}_i$ is the difference between $i$'s pretest score and the classroom mean.
This decomposition was motivated by the possibility that reassignment patterns may differ between high- and low-achieving classrooms, and that a teacher's decision to reassign a student depends on the student's ability relative to the classroom than his or her absolute ability.
Second, we modeled the effect of $\tilde{w}$ on $Z$ as linear in the logit scale, but allowed the slope to vary by classroom.
This was motivated by the possibility that some teachers use reassignment to help struggling students catch up to their peers, so lower $\tilde{w}$ would predict $Z$, and other teachers use it to help high-achievers skip sections related to basic skills, so higher $\tilde{w}$ would predict $Z$.
We also considered models incorporating non-linear effects of
$\tilde{w}$, via natural splines \cite{hastie2017generalized} but
found no evidence that the non-linearity improved the model fit.
We fit the model using the \texttt{lme4} package in \texttt{R}
\cite{lme4}.

@@

% \subsection{Classroom-Level}
% Using data from classrooms in which either all students or no students
% were reassigned, we estimated classroom-level propensity scores
% $Pr(\bar{Z}=1|\bm{\bar{x}})$.
% To do so, we used a single-level logistic regression model:
% \begin{equation}\label{eq:propmod2}
% \begin{split}
% Pr(&\bar{Z}_j=1|\bm{\bar{x}}_j)=\\
% &logit^{-1}( \gamma_{0state[j]}+\gamma_1 \bar{w}_{j}+\\
% &\gamma_2 \%Freshman_j+\gamma_{3} \%FRL_j+\\
% &\gamma_{4}year_j+\gamma_5Nstud_j)
% \end{split}
% \end{equation}
% where $\%Freshman_j$ and $\% FRL_j$ denote the percentages of
% classroom $j$ that are 9th graders or receive free or reduced-price
% lunches, and $Nstud_j$ is the number of students in classroom $j$ in
% our dataset.
% Model \eqref{eq:propmod2} was simpler than the student-level version,
% \eqref{eq:propmod1}, both because it used classroom-aggregated data,
% and because the sample size was smaller ($n=\Sexpr{nclass1+nclass0}$
% classrooms versus $n=nStud1$ students), requiring greater parsimony.
% As described in Section \ref{sec:psm}, model \eqref{eq:propmod2} was
% selected in order to acheive covariate balance \eqref{eq:balance}.

% \begin{figure}\label{fig:coefplot2}
% <<coefplot2,fig.width=3,fig.height=2>>=

% pdat2 <- psmod2coef%>%
%   as.data.frame()%>%
%   rownames_to_column('Covariate')%>%
%   filter(Covariate%in%c('nstud','mpretest','raceBlack','raceHispanic','sexM','grade9','spec_speced', 'spec_gifted','spec_esl','frl','frlMISTRUE','year'))%>%
%   mutate(
%     Covariate= map_chr(Covariate,
%       ~switch(.,
%         mpretest='wbar',#"$\\bar{w}$",
%         pretestC='wtilde',#"$\\tilde{w}$",
%         raceWhite='White',
%         raceBlack='Black',
%         raceHispanic='Latinx',
%         sexF='F',
%         sexM='Male',
%         "grade10+"="10+",
%         grade9="%Freshman",
%         spec_speced="Special Ed.",
%         spec_gifted="Gifted",
%         spec_esl="ESL",
%         frl="%FRL",
%         frlMISTRUE="FRL Missing",
%         year="Year",
%         nstud="Nstud")
%     ),
%     Covariate=factor(Covariate,levels=rev(unique(Covariate))),
%     ymax=Estimate+2*`Std. Error`,
%     ymin=Estimate-2*`Std. Error`
%   )

% ggplot(pdat2,aes(Covariate,Estimate,ymin=ymin,ymax=ymax))+
%       geom_point(size=3)+
%           xlab(NULL)+ylab(NULL)+
%   geom_errorbar(width=0,size=2)+geom_hline(yintercept=0)+
%   scale_x_discrete(labels=c(levels(pdat2$Covariate)[1:(nlevels(pdat2$Covariate)-1)],bquote(bar(w))))+
%   theme(text = element_text(size=10))+
%   coord_flip()

% @
% \caption{Estimated coefficients and 95\% confidence intervals from
%   model \eqref{eq:propmod2}.}
% \label{fig:propmodel2}
% \end{figure}

% Figure \ref{fig:propmodel2} shows estimates and 95\% confidence
% intervals for covariate coefficients from \eqref{eq:propmodel2}.
% Similar to the results from the student-level analysis, reassignment
% is more prevalent in classrooms with lower average pretest scores and
% in the second year of implementation, though \eqref{eq:propmodel2}
% cannot rule out null, or very small opposite-signed effects in either case.
% Teachers in classrooms with a greater proportion of FRL students were less likely
% to use reassignment.

@@

\section{Matching and Covariate Balance}\label{sec:match}
%\subsection{Student Level Match}
We construct a student-level match based on propensity scores on the
log-odds scale, i.e. $log(\hat{e}/(1-\hat{e})$.
Instead of a pair-matching design, which would necessitate discarding
non-reassigned students who would make good matched comparisons, we
use a restricted full match design \cite{hansen2004}.
In this design, the numbers of reassigned and not-reassigned students
in each matched set is allowed to vary, so that in some cases several
reassigned students may be matched with a single non-reassigned
student, and vice-versa.
We use the \texttt{R} package \texttt{optmatch} \cite{optmatch} to
choose the matched sets optimally.
The \texttt{fullmatch()} routine takes a matrix of discrepancies
(e.g. differences in propensity scores) between treatment and control
subjects, and arranges them into matched sets so that the sum of
absolute discrepancies between matched subjects is minimized.

As described at the end of Section \ref{sec:psm}, the post-test scores
played no role in this process.
Hence, we were able to iteratively match students, check covariate
balance, modify the propensity score model and/or the matching routine
if necessary, and repeat until adequate balance was achieved.
Here we present the final match; a record of attempts is available on
the first author's github site.

\begin{figure}
\includegraphics[width=0.95\linewidth]{plots/psPlot1.pdf}
\caption{Estimated propensity scores for reassigned and not-reassigned
  students. Scores for students who were excluded from the ultimate
  match are colored red.}
\label{fig:psplot1}
\end{figure}
